---
title: "exercise-12"
author: "Erika Carlson"
date: 2024-04-14
format: html
editor_options: 
  chunk_output_type: console
---

# Exercise 12 {.unnumbered}

# Practice Model Selection {.unnumbered}

## Preliminaries {.unnumbered}

* Load the "Mammal_lifehistories_v2" dataset as a "tibble" named **d**. 
  + This dataset compiles life history and other variables for over 1400 species of placental mammals from 17 different Orders.

> **Data source**:
>
> Ernest SKM. (2003). Life history characteristics of placental nonvolant mammals. *Ecology* 84: 3402–3402.

```{r}
library(tidyverse)
d <- read_tsv("https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/Mammal_lifehistories_v2.txt", col_names = TRUE)
```

* Use {skimr} package to conduct exploratory data analysis. 

```{r}
library(skimr)
skim(d)
```

* Which of the variables are categorical and which are numeric?
  + *Categorical: order, family, Genus, and species* 
  + *Numeric: mass(g), gestation(mo), newborn(g), weaning(mo), wean mass(g), AFR(mo), max. life(mo), litter size, litters/year, refs*

## Challenge {.unnumbered}
#### Step 1 {.unnumbered}

* Replace all values of -999 (the authors' code for missing data) with `NA`.

```{r}
d <- d %>% mutate(across(`mass(g)`:`litters/year`, \(x) na_if(x, -999)))
# For `across()` Supply arguments directly to `.fns` through an anonymous function
```

#### Step 2 {.unnumbered}

* Drop the variables **litter size** and **refs**.

```{r}
d <- d %>% dplyr::select(-refs, -`litter size`)
```

#### Step 3 {.unnumbered}

* Log transform **all** of the other numeric variables.

```{r}
s <- d %>% mutate(across(`mass(g)`:`litters/year`, \(x) log(x)))

# make new column names
s <- s %>% rename(logMass = `mass(g)`,
                  logGest = `gestation(mo)`,
                  logNewbornMass = `newborn(g)`,
                  logWean = `weaning(mo)`,
                  logWeaningMass = `wean mass(g)`,
                  logAFR = `AFR(mo)`,
                  logLife = `max. life(mo)`,
                  logLitter = `litters/year`
                  )
```

#### Step 4 {.unnumbered}

* Regress the (now log transformed) age [**logGest**, **logWean**, **logAFR** (age at first reproduction), and **logLife** (maximum lifespan)] and mass [**logNewbornMass** and **logWeaningMass**] variables on (now log transformed) overall body **logMass** and add the residuals to the dataframe as new variables (**relGest**, **relWean**, **relAFR**, **relLife**, **relNewbornMass**, and **relWeaningMass**).

> **HINT:** Use "na.action=na.exclude" in your`lm()` calls. With this argument set, the residuals will be padded to the correct length by inserting NAs for cases with missing data.

```{r}
relGest_model <- lm(logMass ~ logGest, na.action = na.exclude, data = s)
s$relGest <- resid(relGest_model)

relWean_model <- lm(logMass ~ logWean, na.action = na.exclude, data = s)
s$relWean <- resid(relWean_model)

relAFR_model <- lm(logMass ~ logAFR, na.action = na.exclude, data = s)
s$relAFR <- resid(relAFR_model)

relLife_model <- lm(logMass ~ logLife, na.action = na.exclude, data = s)
s$relLife <- resid(relLife_model)

relNewbornMass_model <- lm(logMass ~ logNewbornMass, na.action = na.exclude , data = s)
s$relNewbornMass <- resid(relNewbornMass_model)

relWeaningMass_model <- lm(logMass ~ logWeaningMass, na.action = na.exclude, data = s)
s$relWeaningMass <- resid(relWeaningMass_model)
```

#### Step 5 {.unnumbered}

* Plot residuals of max lifespan (**relLife**) in relation to **Order**. Which mammalian orders have the highest residual lifespan?
  + *Artiodactyla, Cetacea, Perissodactyla, Proboscidea*
* Plot residuals of newborn mass (**relNewbornMass**) in relation to **Order**. Which mammalian orders have the have highest residual newborn mass?
  + *Carnivora, Perissodactyla*
  + *Proboscidea, Tubulidentata also high but very low n*
* Plot residuals of weaning mass (**relWeaningMass**) in relation to **Order**. Which mammalian orders have the have highest relative weaning mass?
  + *Pholidota, Lagomorpha, Carnivora *
  + *Sirenia, Tubulidentata, Xenarthra, Hyracoidea also high but very low n*

> **NOTE:** There will be lots of missing data for the latter two variables!

```{r}
p1 <- ggplot(data = s %>% drop_na(relLife),
             aes(x = order, y = relLife)) +
  geom_boxplot() +
  geom_point(alpha = 0.4) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  scale_y_continuous(limits = c(NA, NA))
p1

p2 <- ggplot(data = s %>% drop_na(relNewbornMass),
             aes(x = order, y = relNewbornMass)) +
  geom_boxplot() +
  geom_point(alpha = 0.4) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  scale_y_continuous(limits = c(NA, NA))
p2

p3 <- ggplot(data = s %>% drop_na(relWeaningMass),
             aes(x = order, y = relWeaningMass)) +
  geom_boxplot() +
  geom_point(alpha = 0.4) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  scale_y_continuous(limits = c(NA, NA)) 
p3
```

#### Step 6 {.unnumbered}

* Run models and a model selection process to evaluate what variables best predict each of the two response variables, **logLife** and **logAFR**, from the set of the following predictors: **logGest**, **logNewbornMass**, **logWean**, **logWeaningMass**, **logLitters**, and **logMass**.

```{r}
# for logLife

# winnow dataset to drop rows that are missing the response variable or any of the predictors
s1 <- s %>% drop_na(logLife, logGest, logNewbornMass, logWeaningMass, logWean, logLitter, logMass)

# model selection using reverse screening (the long way)
full <- lm(data = s1, logLife ~ logGest + logNewbornMass + logWeaningMass + logWean + logLitter + logMass)
summary(full) 

drop1(full, test = "F") # logWeaningMass has non-sig p-value, lowest F, and lowest increase in RSS 

m1 <- update(full, . ~ . - logWeaningMass)
summary(m1)

drop1(m1, test = "F") # logNewbornMass has non-sig p-value, lowest F, and lowest increase in RSS

m2 <- update(m1, . ~ . - logNewbornMass)
summary(m2)

drop1(m2, test = "F") # all explanatory variables are significant, model 2 is best

# logWean has lowest F value and lowest increase in RSS
m3 <- update(m2, . ~ . - logWean)
summary(m3) # lower adjusted R-squared than model 2

# model selection using AIC (to verify)
library(MASS)
(step1 <- stepAIC(full_m1, scope = . ~ ., direction = "both"))
detach(package:MASS)

# model comparisons using AICc
library(MuMIn)
full <- lm(data = s1, logLife ~ logGest + logNewbornMass + logWeaningMass + logWean + logLitter + logMass, na.action = na.fail)
# include 'na.fail' in 'na.action' argument to prevent `dredge()` from trying to fit submodels with different data sets than the global model if there are missing values

mods <- dredge(full)
(mods_delta <- subset(mods, delta <= 4)) # limit models to delta AICc of 4 of less

# average coefficients for top models where delta.aicc <= 4
mods.avg <- summary(model.avg(mods, subset = delta <= 4, fit = TRUE))

mods.avg$msTable

# CIs for averaged beta coefficient estimates
confint(mods.avg)

plot(mods.avg, full = TRUE)

detach(package:MuMIn)
```

* The best model for **logLife** includes predictor variables **logWean**, **logLitter**, **logGest**, and **logMass** with AICc of 264.6
* Five models have a delta AICc of 4 or less
* **logGest**, **logLitter**, and **logWean** appear in all "top" models

```{r}
# for logAFR

# winnow dataset to drop rows that are missing the response variable or any of the predictors
s2 <- s %>% drop_na(logAFR, logGest, logNewbornMass, logWeaningMass, logWean, logLitter, logMass)

# model selection using reverse screening (the long way)
full <- lm(data = s2, logAFR ~ logGest + logNewbornMass + logWeaningMass + logWean + logLitter + logMass)
summary(full) 

drop1(full, test = "F") # logWeaningMass has non-sig p-value, lowest F, and lowest increase in RSS 

m1 <- update(full, . ~ . - logWeaningMass)
summary(m1)

drop1(m1, test = "F") # logNewbornMass has non-sig p-value, lowest F, and lowest increase in RSS

m2 <- update(m1, . ~ . - logNewbornMass)
summary(m2)

drop1(m2, test = "F") # all explanatory variables are significant, model 2 is best

# logWean has lowest F value and lowest increase in RSS
m3 <- update(m2, . ~ . - logWean)
summary(m3) # lower adjusted R-squared than model 2

# model selection using AIC (to verify)
library(MASS)
(step1 <- stepAIC(full, scope = . ~ ., direction = "both"))
detach(package:MASS)

# model comparisons using AICc
library(MuMIn)
full <- lm(data = s2, logAFR ~ logGest + logNewbornMass + logWeaningMass + logWean + logLitter + logMass, na.action = na.fail)
# include 'na.fail' in 'na.action' argument to prevent `dredge()` from trying to fit submodels with different data sets than the global model if there are missing values

mods <- dredge(full)
(mods_delta <- subset(mods, delta <= 4)) # limit models to delta AICc of 4 of less

# average coefficients for top models where delta.aicc <= 4
mods.avg <- summary(model.avg(mods, subset = delta <= 4, fit = TRUE))

mods.avg$msTable

# CIs for averaged beta coefficient estimates
confint(mods.avg)

plot(mods.avg, full = TRUE)

detach(package:MuMIn)
```

* The best model for **logAFR** includes predictor variables **logWean**, **logLitter**, **logGest**, and **logMass** with AICc of 412.0
* Seven models have a delta AICc of 4 or less
* **logGest**, **logLitter**, and **logWean** appear in all "top" models
